### 2020-02-20 thesis/content/chapter-related-work.tex
continued wl section

\begin{defn}
	A graph $G \coloneqq (\mathcal{V}_G, \mathcal{E}_G)$ consists of a set of vertices $v_i \in \mathcal{V}_G$ and a set of edges $e_{ij} = (v_i, v_j) \in \mathcal{E}_G \subseteq \mathcal{V}_G^2$.
	Optionally discrete vertex labels $l_i \in L_{\mathcal{V}}$ or edge labels $l_{ij} \in L_{\mathcal{E}}$ may be associated with all vertices $v_i \in \mathcal{V}_G$ and edges $e_{ij}\in \mathcal{E}_G$ respectively.
	Also continuous feature vectors $x_{i} \in \mathcal{X}_{\mathcal{V}}$, $x_{ij} \in \mathcal{X}_{\mathcal{E}}$ may be given.
	If $\mathcal{X}_{\mathcal{E}} = \mathbb{R}$, $x_{ij}$ can for example be interpreted as an edge weight of $e_{ij}$.
\end{defn}
In this thesis only undirected graphs $G$ will be considered, i.e.\ ${e_{ij} \in \mathcal{E}_G} \leftrightarrow\allowbreak {e_{ji} \in \mathcal{E}_G} \allowbreak\land l_{ij} = l_{ji} \allowbreak\land x_{ij} = x_{ji}$.
To classify or score such a graph, it first needs to be characterized by a set of relevant properties.
The most strict characterization of a graph is its so-called \textit{isomorphism type}.
It uniquely identifies a graph but lacks any notion of similarity between non-identical graphs.
We will begin with a brief definition of this strict isomorphism-based graph characterization.
Then two less strict characterization approaches are described; the so called \acl{wl} coloring and the notion of graph spectra.
They are the theoretical foundation of many current \ac{gcr} methods.

@@

@@

The \acf{wl} algorithm~\cite{Weisfeiler1968}\cite{Cai1992} characterizes a graph $G$ by assigning discrete labels $c \in \mathcal{C}$, called \textit{colors}, to vertex $k$-tuples $(v_1, \dots, v_k) \in \mathcal{V}_G^k$, where $k \in \mathbb{N}$ is the freely choosable \textit{\ac{wl}-dimension}.
A mapping $\chi_{G, k}: \mathcal{V}_G^k \to \mathcal{C}$ is called a \textit{$k$-coloring} of $G$.
\begin{defn}
	A coloring $\chi'$ \textit{refines} $\chi$ ($\chi' \succeq \chi$) iff.\ $\forall a, b \in \mathcal{V}_G^k: \chi(a) \neq \chi(b) \rightarrow \chi'(a) \neq \chi'(b)$, i.e.\ $\chi'$ distinguishes at least those tuples that are distinguished by $\chi$.
\end{defn}
\begin{defn}
	Two colorings $\chi$ and $\chi'$ are \textit{equivalent} ($\chi \equiv \chi'$) iff.\ $\forall a, b \in \mathcal{V}_G^k: {\chi(a) = \chi(b)} \leftrightarrow {\chi'(a) = \chi'(b)}$, i.e.\ $\chi$ is identical to $\chi'$ up to color substitutions.
\end{defn}
The $k$-dimensional \ac{wl} algorithm ($k$-\acs{wl}) works by iteratively refining $k$-colorings $\chi_{G, k}^{(0)} \preceq \chi_{G, k}^{(1)} \preceq \dots$ of a given graph $G$ until the convergence criterion $\chi_{G, k}^{(i)} \equiv \chi_{G, k}^{(i+1)}$ is satisfied.
We denote the final, maximally refined, $k$-\acs{wl} coloring with $\chi^{*}_{G, k}$.
\begin{defn}\label{defn:related:wl-distinguishable}
	Two graphs $G$ and $H$ are are $k$-\acs{wl} \textit{distinguishable} ($G \mathrel{{\not\simeq}_k} H$) iff.\ there exists a color $c \in \mathcal{C}$ s.t.\ $\left|\left\{ a \in \mathcal{V}_G^k\, |\, \chi^{*}_{G, k}(a) = c \right\}\right|\neq \left|\left\{ b \in \mathcal{V}_H^k\, |\, \chi^{*}_{H, k}(b) = c \right\}\right|$. % chktex 21
\end{defn}
It is easy to show that $G \mathrel{{\not\simeq}_k} H \implies G \not\simeq H$.
The opposite does however not necessarily hold;
two $k$-\acs{wl} indistinguishable graphs are not always isomorphic, i.e.\ $G \mathrel{{\simeq}_k} H \centernot\implies G \simeq H$.
To understand why this is the case and which types of graphs are $k$-\acs{wl} distinguishable, we will now look at the details of \ac{wl} color refinement step.
First the color refinement algorithm for the most simple case of $k = 1$ is described.
Then the definitions and intuitions from the 1-dimensional case are extended to its higher-dimensional generalization.
Lastly we will discuss the discriminative power of the \acs{wl} algorithm and its relation to the \acs{wl}-dimension $k$.

\subsubsection{The 1-dimensional \acs{wl} algorithm}
In the 1-dimensional \ac{wl} algorithm, a color is assigned to each vertex of a graph.
If the vertices $\mathcal{V}_G$ of the input graph $G$ are already labeled via some labeling $l_G: \mathcal{V}_G \to L$, those labels can be used as the initial graph coloring $\chi_{G,1}^{(0)} = l_G$.
For unlabeled graphs a constant coloring is used instead, e.g.\ $\forall v \in \mathcal{V}_G: \chi_{G,1}^{(0)}(v) = \textcolor{t_blue}{\texttt{A}}$ for some initial color $\textcolor{t_blue}{\texttt{A}} \in \mathcal{C}$. % chktex 25
In each iteration of the 1-\acs{wl} color refinement algorithm, the following neighborhood aggregation scheme is used to compute a new color for all vertices:
\begin{defn}\label{defn:related:wl1-refine}
	$\chi_{G,1}^{(i+1)}(v) \coloneqq h\left(\chi_{G,1}^{(i)}(v), \ldblbrace \chi_{G,1}^{(i)}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right)$,
	with $\Gamma_G(v)$ denoting the set of adjecent vertices of $v \in \mathcal{V}_G$ and $h: \mathcal{C}^* \to \mathcal{C}$ denoting a perfect hash function that assigns a unique color to each finite combination of colors.
	In practice $h$ is usually defined lazily by using $\mathcal{C} = \mathbb{N}$ and non-deterministically enumerating color combinations in the order in which they are hashed s.t.\ a new color is introduced every time a previously unseen color combination is queried.
\end{defn}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.75\linewidth]{gfx/related-work/wl1-refine.pdf}
	\caption[Example 1-WL color refinement steps.]{
		Example 1-WL color refinement steps.
		After two iterations the coloring stabilizes.
		Each vertex is labeled with its current color and has the aggregated neighbors, as in \cref{defn:related:wl1-refine}, written next to it.
	}\label{fig:related:wl1-refine}
\end{figure}

@@

Let us now look at how the tuple colors are initialized.
For this we use subgraph \textit{isomorphism types} which determine the initial color $\chi_{G,k}^{(0)}(v)$ of each $k$-tuple $v$.
For $k = 1$ the isomorphism type of a vertex $v$ directly corresponds to its label $l_G(v)$.
For $k > 1$ the isomorphism types more generally correspond to the classes of isomorphic ordered subgraphs of $G$ induced by the $k$-tuples $\mathcal{V}_G^k$.
Formally this means that
\begin{align}
	\chi_{G,k}^{(0)}(v_1, \dots, v_k) = \chi_{G,k}^{(0)}(w_1, \dots, w_k) \label{eq:related:wlk-init}
	\iff &\ \forall i: l_G(v_i) = l_G(w_i) \\
	\land &\ \forall i, j: v_i = v_j \leftrightarrow w_i = w_j \nonumber \\
	\land &\ \forall i, j: (v_i, v_j) \in \mathcal{E}_G \leftrightarrow (w_i, w_j) \in \mathcal{E}_G \text{.} \nonumber
\end{align}
If no explicit vertex labeling $l_G$ is given, a single constant label for all vertices is assumed.
Also note that there is a fundamental difference in how the adjacency information encoded in $\mathcal{E}_G$ is used in 1-\acs{wl} vs.\ $k$-\acs{wl}:
In 1-\acs{wl} a vertex coloring by itself cannot encode adjacency which is why this information is explicitly incorporated in each refinement step via $\Gamma_G$ (see~\cref{defn:related:wl1-refine}).
In $k$-\acs{wl} on the other hand each pair of vertices $(v, u) \in \mathcal{V}_G^2$ appears in at least one $k$-tuple (assuming $k \geq 2$) and therefore has at least one color which can implicitly encode the adjacency information.
Edges and non-edges are colored differently in the initial coloring as defined in \cref{eq:related:wlk-init};
thus no explicit adjacency information is needed in the $k$-\acs{wl} color refinement step in \cref{defn:related:wlk-refine}.

>>>>

\begin{defn}
	A graph $G \coloneqq (\mathcal{V}_G, \mathcal{E}_G)$ consists of a set of vertices $v_i \in \mathcal{V}_G$ and a set of edges $e_{ij} = (v_i, v_j) \in \mathcal{E}_G \subseteq \mathcal{V}_G^2$.
	Optionally discrete vertex labels $l_G[v_i] \in L_{\mathcal{V}}$ or edge labels $l_G[e_{ij}] \in L_{\mathcal{E}}$ may be associated with all vertices $v_i \in \mathcal{V}_G$ and edges $e_{ij}\in \mathcal{E}_G$ respectively.
	Also continuous feature vectors $x_G[v_{i}] \in \mathcal{X}_{\mathcal{V}}$, $x_G[e_{ij}] \in \mathcal{X}_{\mathcal{E}}$ may be given.
	If $\mathcal{X}_{\mathcal{E}} = \mathbb{R}$, $x_G[e_{ij}]$ can for example be interpreted as an edge weight of $e_{ij}$.
\end{defn}
In this thesis all graphs $G$ are assumed to be undirected if not explicitly stated otherwise, i.e.\ ${e_{ij} \in \mathcal{E}_G} \leftrightarrow\allowbreak {e_{ji} \in \mathcal{E}_G} \allowbreak\land l_G[e_{ij}] = l_G[e_{ji}] \allowbreak\land x_G[e_{ij}] = x_G[e_{ji}]$.
To classify or score such a graph, it first needs to be characterized by a set of relevant properties.
The most strict characterization of a graph is its so-called \textit{isomorphism type}.
It uniquely identifies a graph but lacks any notion of similarity between non-identical graphs.
We will begin with a brief definition of this strict isomorphism-based graph characterization.
Then two less strict characterization approaches are described; the so called \acl{wl} coloring and the notion of graph spectra.
They are the theoretical foundation of many current \ac{gcr} methods.

@@

In order to process a graph $G$ of size $n = |\mathcal{V}_G|$ , one is forced to choose some encoding, e.g.\ an adjacency matrix $A \in {\{0, 1\}}^{n \times n}$.
Such an encoding introduces a vertex ordering $v_1, \dots, v_n$ that does not carry any semantic meaning.
This means that there are $n!$ equivalent encodings of $G$. % chktex 40
\begin{defn}
	Two encoded graphs $G$ and $H$ with vertices $\mathcal{V}_G = \{ v_1, \dots, v_n \}$ and $\mathcal{V}_H = \{ w_1, \dots, w_n \}$ respectively are \textit{isomorphic} ($G \simeq H$) iff.\
\end{defn}

@@

The \acf{wl} algorithm~\cite{Weisfeiler1968}\cite{Cai1992} characterizes a graph $G$ by assigning discrete labels $c \in \mathcal{C}$, called \textit{colors}, to vertex $k$-tuples $(v_1, \dots, v_k) \in \mathcal{V}_G^k$, where $k \in \mathbb{N}$ is the freely choosable \textit{\ac{wl}-dimension}.
A mapping $\chi_{G, k}: \mathcal{V}_G^k \to \mathcal{C}$ is called a \textit{$k$-coloring} of $G$.
\begin{defn}
	A coloring $\chi'$ \textit{refines} $\chi$ ($\chi' \succeq \chi$) iff.\ $\forall a, b \in \mathcal{V}_G^k: \chi(a) \neq \chi(b) \rightarrow \chi'(a) \neq \chi'(b)$, i.e.\ $\chi'$ distinguishes at least those tuples that are distinguished by $\chi$.
\end{defn}
\begin{defn}
	Two colorings $\chi$ and $\chi'$ are \textit{equivalent} ($\chi \equiv \chi'$) iff.\ $\forall a, b \in \mathcal{V}_G^k: {\chi(a) = \chi(b)} \leftrightarrow {\chi'(a) = \chi'(b)}$, i.e.\ $\chi$ is identical to $\chi'$ up to color substitutions.
\end{defn}
The $k$-dimensional \ac{wl} algorithm ($k$-\acs{wl}) works by iteratively refining $k$-colorings $\chi_{G, k}^{(0)} \preceq \chi_{G, k}^{(1)} \preceq \dots$ of a given graph $G$ until the convergence criterion $\chi_{G, k}^{(i)} \equiv \chi_{G, k}^{(i+1)}$ is satisfied.
We denote the final, maximally refined, $k$-\acs{wl} coloring with $\chi^{*}_{G, k}$.
\begin{defn}\label{defn:related:wl-distinguishable}
	Two graphs $G$ and $H$ are are $k$-\acs{wl} \textit{distinguishable} ($G \mathrel{{\not\simeq}_k} H$) iff.\ there exists a color $c \in \mathcal{C}$ s.t.\ $\left|\left\{ a \in \mathcal{V}_G^k\, |\, \chi^{*}_{G, k}(a) = c \right\}\right|\neq \left|\left\{ b \in \mathcal{V}_H^k\, |\, \chi^{*}_{H, k}(b) = c \right\}\right|$. % chktex 21
\end{defn}
As we will see, the \ac{wl} color refinement does not depend on vertex or edge encoding orderings, thus any difference in the final coloring of two graphs always implies the non-isomorphism of the colored graphs, i.e.\ $G \mathrel{{\not\simeq}_k} H \implies G \not\simeq H$.
The opposite does however not necessarily hold;
two $k$-\acs{wl} indistinguishable graphs are not always isomorphic, i.e.\ $G \mathrel{{\simeq}_k} H \Longarrownot\implies G \simeq H$.
To understand why this is the case and which types of graphs are $k$-\acs{wl} distinguishable, we will now look at the details of \ac{wl} color refinement step.
First the color refinement algorithm for the most simple case of $k = 1$ is described.
Then the definitions and intuitions from the 1-dimensional case are extended to its higher-dimensional generalization.
Lastly we will discuss the discriminative power of the \acs{wl} algorithm and its relation to the \acs{wl}-dimension $k$.

\subsubsection{The 1-dimensional \acs{wl} algorithm}
In the 1-dimensional \ac{wl} algorithm, a color is assigned to each vertex of a graph.
If the vertices $v \in \mathcal{V}_G$ of the input graph $G$ are labeled, those labels $l_G[v] \in L_{\mathcal{V}} \subseteq \mathcal{C}$ can be used as the initial graph coloring $\chi_{G,1}^{(0)}(v) \coloneqq l_G[v]$.
For unlabeled graphs a constant coloring is used instead, e.g.\ $\forall v \in \mathcal{V}_G: \chi_{G,1}^{(0)}(v) = \textcolor{t_blue}{\texttt{A}}$ for some initial color $\textcolor{t_blue}{\texttt{A}} \in \mathcal{C}$. % chktex 25
In each iteration of the 1-\acs{wl} color refinement algorithm, the following neighborhood aggregation scheme is used to compute a new color for all vertices:
\begin{defn}\label{defn:related:wl1-refine}
	$\chi_{G,1}^{(i+1)}(v) \coloneqq h\left(\chi_{G,1}^{(i)}(v), \ldblbrace \chi_{G,1}^{(i)}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right)$,
	with $\Gamma_G(v)$ denoting the set of adjecent vertices of $v \in \mathcal{V}_G$ and $h: \mathcal{C}^* \to \mathcal{C}$ denoting a perfect hash function that assigns a unique color to each finite combination of colors.
\end{defn}
In practice the hash function $h$ is usually defined lazily by using $\mathcal{C} = \mathbb{N}$ and non-deterministically enumerating color combinations in the order in which they are hashed s.t.\ a new color is introduced every time a previously unseen color combination is queried.
This is illustrated in \cref{fig:related:wl1-refine}.
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.75\linewidth]{gfx/related-work/wl1-refine.pdf}
	\caption[Example 1-WL color refinement steps.]{
		Example 1-WL color refinement steps.
		After two iterations the coloring stabilizes.
		Each vertex $v$ is labeled with its current color and has its previous color and the colors of the hashed neighbors $\Gamma_G(v)$ written next to it (see \cref{defn:related:wl1-refine}).
	}\label{fig:related:wl1-refine}
\end{figure}

@@

Let us now look at how the tuple colors are initialized.
For this we use subgraph \textit{isomorphism types} which determine the initial color $\chi_{G,k}^{(0)}(v)$ of each $k$-tuple $v$.
For $k = 1$ the isomorphism type of a vertex $v$ directly corresponds to its label $l_G[v]$.
For $k > 1$ the isomorphism types more generally correspond to the classes of isomorphic ordered subgraphs of $G$ induced by the $k$-tuples $\mathcal{V}_G^k$.
Formally this means that
\begin{align}
	\chi_{G,k}^{(0)}(v_1, \dots, v_k) = \chi_{G,k}^{(0)}(w_1, \dots, w_k) \label{eq:related:wlk-init}
	\iff &\ \forall i: l_G[v_i] = l_G[w_i] \\
	\land &\ \forall i, j: v_i = v_j \leftrightarrow w_i = w_j \nonumber \\
	\land &\ \forall i, j: (v_i, v_j) \in \mathcal{E}_G \leftrightarrow (w_i, w_j) \in \mathcal{E}_G \text{.} \nonumber
\end{align}
If no explicit vertex labeling $l_G$ is given, a single constant label for all vertices is assumed.
Also note that there is a fundamental difference in how the adjacency information encoded in $\mathcal{E}_G$ is used in 1-\acs{wl} vs.\ $k$-\acs{wl}:
In 1-\acs{wl} a vertex coloring by itself cannot encode adjacency which is why this information is explicitly incorporated in each refinement step via $\Gamma_G$ (see~\cref{defn:related:wl1-refine}).
In $k$-\acs{wl} on the other hand each pair of vertices $(v, u) \in \mathcal{V}_G^2$ appears in at least one $k$-tuple (assuming $k \geq 2$) and therefore has at least one color which can implicitly encode the adjacency information.
Edges and non-edges are colored differently in the initial coloring as defined in \cref{eq:related:wlk-init};
thus no explicit adjacency information is needed in the $k$-\acs{wl} color refinement step in \cref{defn:related:wlk-refine}.
